---
title: "Session 2: Data Manipulation & Visualization -- Solutions"
author: 'Written by Amy Campbell, -, -, and - in direct reference to exercises developed by Shweta Ramdas, Binglan Li, and Matt Paul for PennPrep 2019'
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading Installed Packages
```{r, warning=FALSE,message=FALSE,error=FALSE}
library("dplyr")
library("ggplot2")
library("plyr")
```

## Loading the Parkinsons' Disease gene expression data (Courtesy of Shweta Ramdas )
Let's first have a look at our data. 
```{r}
# Load in Parkinsons Disease dataset 
# Note: sep='\t' specifies tab-delimited file, and stringsAsFactors tells 
# R to encode categorical variables as 'factors' -- more on that later. 
expression = read.table("expressionpd.txt", sep = '\t', stringsAsFactors=TRUE)
# How big is it? 
print(dim(expression))
# dimensions: [10  19101] -- 10 rows and 19101 columns
```
With 19101 variables, we should start by taking a look at the first few column names rather than hitting 'View()'. 

```{r}
print(head(colnames(expression)))
```
Let's take a look at the two categorical variables built into the dataset:
```{r}
# Encodes whether a sample is from a person with/without Parkinsons' Disease
print(head(expression$samples))
# Encodes whether a sample is from a male/female individual 
print(head(expression$sex))
# How many observations do we have of each variable? 
print(plyr::count(expression, vars=c("sex", "samples")))
```
## Basic Visualization
Let's use PARK7 as an example of a gene with known associations with Parkinsons' disease. Let's compare this gene's expression level in control subjects vs. subjects with Parkinsons':
```{r}
ggplot2::ggplot(expression, aes(x= samples, y=PARK7, fill=samples)) + geom_boxplot() + labs(title="PARK7 Gene Expression in Parkinsons Disease Samples vs. Controls")
```
