---
title: "SOLUTION: Practicum for Session 2 of R workshop for Incoming GCB Students (2019)"
subtitle: "By: Amy Campbell"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

## Introduction
In this practicum, we'll put the data manipulation and visualization practices we've talked about to use. Things we want you to try to do are written in <span style="color:red">**red**</span>. 


Functions that will prove useful:
**base R **
  * summary()
**ggplot2**

  * ggplot()
  * facet_grid()
  * ggtitle()
  * scale_color_manual()
  * scale_fill_manual()
  * qplot()
  * geom_boxplot()
  * geom_histogram()
  * geom_bar()

**dplyr**

  * group_by()
  * mutate()
  * filter()
  * %>% 
  * arrange()
  * ifelse()

Remember, you can access one variable in a dataframe using **$**. 

## The dataset

The data we'll be using are from city-provided reports of energy & natural resource usage by large buildings in Philadelphia, accessed via https://www.opendataphilly.org/dataset/large-commercial-building-energy-benchmarking . For simplicity, we've subsetted the data to 10 zip codes (for full data pre-processing, you can (optionally) look at github:gcb-r-workshop/session2/Practicum/formatting_energy.R) 

It's a comma-separated text file, and its filename is *city_energy_data.csv*. Rows represent properties and columns represent different variables. The first row of the file contains the variable names ('header')

<span style="color:red">**Read the file in and save it as a variable called 'energy'**</span>
```{r}
energy = read.csv("/Users/amycampbell/city_energy_data.csv")
```

<span style="color:red">**Print out the names of the variables, and see if you can get a sense of the distributions/ranges of the variables 'year_built' and  'squarefootage' **</span>


```{r}
# Printing variable names & first few values of each
colnames(energy)
head(energy)
# Get quartiles for 3 variables
summary(energy$year_built)
summary(energy$squarefootage)
```

That's a pretty big spread of square footages, huh? 

<span style="color:red">**Try to use ggplot2 to make a histogram of square footage. **</span>

 Hint: If it's hard to see the distribution with a histogram of raw values, try using a transformation like log() log10() -- Make sure to include a title! 
```{r}

ggplot2::ggplot(energy, aes(x=squarefootage)) + geom_histogram(bins=100) 
# log 10 transformed is easier to make sense of
ggplot2::ggplot(energy, aes(x=log10(squarefootage))) + geom_histogram(bins=100) + ggtitle("Histogram of Log-transformed Building Square Footage")

```

Note that if you just use the default geom_histogram() call, you'll get a gray background with darker grey bars. If you want, <span style="color:red">**try selecting the fill color by looking up color keys, or by setting a palette using RColorBrewer. While you're at it, make a similar histogram of greenhouse_emissions.**</span>  Here's a list of RColorBrewer palettes. You can get an ordered list of values corresponding to the first 'n' colors in one of these palettes using RColorBrewer::brewer.pal( "<palette_name>"). In geom_histogram(), 'color' sets the color of the bar outlines, and 'fill' sets the color filling the bars. As long as 
```{r, fig.height=7}
RColorBrewer::display.brewer.all()
```
```{r}
ggplot2::ggplot(energy, aes(x=log10(squarefootage))) + geom_histogram(bins=100, fill=RColorBrewer::brewer.pal(3, "Dark2")[2], color="black") + ggtitle("Histogram of Log-transformed Building Square Footage") + scale_fill_brewer(palette="Dark2")+ theme(plot.title = element_text(size=14, face="bold"))
```


It's not normally distributed, but it seems like log-scaled axes will help us see any patterns in this variable more easily. <span style="color:red">**Next, make the same plots but for the greenhouse_emissions variable.**</span>
```{r}

ggplot2::ggplot(energy, aes(x=(greenhouse_emissions))) + geom_histogram(bins=100, fill=RColorBrewer::brewer.pal(3, "Set1")[3], color="black") + ggtitle("Histogram of  Greenhouse Gas Emissions") + scale_fill_brewer(palette="Set1")+ theme(plot.title = element_text(size=14, face="bold"))

ggplot2::ggplot(energy, aes(x=log10(greenhouse_emissions))) + geom_histogram(bins=100, fill=RColorBrewer::brewer.pal(3, "Set1")[3], color="black") + ggtitle("Histogram of Log-transformed Greenhouse Gas Emissions") + scale_fill_brewer(palette="Set2")+ theme(plot.title = element_text(size=14, face="bold"))
```

 <span style="color:red">**Next, make a scatterplot of greenhouse_emissions vs. squarefootage.**</span>. It can be a simple one for now, but use ggplot2's scatterplot function geom_point(). Remember what you saw in the above histograms when you're choosing how to plot this. 

```{r}
ggplot2::ggplot(energy, aes(y=log10(greenhouse_emissions), x=log10(squarefootage))) + geom_point() + xlab("Log10-transformed Square Footage") + ylab("Log10-transformed Greenhouse Emissions") + ggtitle("Greenhouse Emissions & Building Square Footage") + theme(plot.title = element_text(size=14, face="bold"))
```

Now, let's consider whether this trend is the same across the years the buildings were constructed. First, we'll make a new factor in the **energy** dataframe coded by the range of years a property was built in. Just so our 'facets' are evenly distributed, we'll divide them into year_built quartiles based on the results of our summary(energy$year_built) call.  **year_built** by  <span style="color:red">**Add a variable to the dataframe called "year_range" that splits year_built into the following groups: 1756-1927 ; 1928-1968; 1969-1994; 1994-2017.**</span>. **Hint**: One way to do this is to use dplyr's **mutate** with the ifelse() statement. You might want to ignore properties without recorded year_built (is.na(year_built)). Save this new dataframe with a new name.

```{r}
# Making this factor using ifelse()
energy_updated =energy %>% filter(!is.na(year_built)) %>% mutate(year_range=(ifelse(year_built<=1927, "1756-1927", ifelse(year_built> 1927 & year_built<= 1968,"1928 - 1969", ifelse(year_built > 1968 & year_built <=1994, "1970-1994", "1995-2017")))))
# Making this factor using case_when() 
energy_updated_casewhen =energy %>% filter(!is.na(year_built)) %>%
  mutate(year_range=case_when(
         (year_built <= 1927) ~ "1756-1927", 
         (year_built > 1927 & year_built<= 1968) ~ "1928 - 1968",
         (year_built > 1968 & year_built <=1994)~ "1969-1994", 
         year_built > 1994 ~"1995-2017")
         
         )
```

Now that we have the **year_range** variable, we can use it to split the previous scatterplot into 4 side-by-side plots of different building periods. <span style="color:red">**Use the year_range variable to plot greenhouse_emissions against squarefootage as 4 side-by-side scatterplots separated by year_built.*</span>. In case you don't have a favorite color palette, here's one you can use with scale_color_manual(values=c()): cbm_palette <- c("#9900FF","#FF9900", "#0072B2", "#33CC33", "#663300", "#CC99FF", "#F0E442", "#4D4D4D", "#CC0000", "#006600", "#D119A3", "#56B4E9")

```{r}

cbm_palette <- c("#9900FF","#FF9900", "#0072B2", "#33CC33", "#663300", "#CC99FF", "#F0E442", "#4D4D4D", "#CC0000", "#006600", "#D119A3", "#56B4E9")

ggplot2::ggplot(energy_updated, aes(y=log10(greenhouse_emissions), x=log10(squarefootage), color=as.factor(property_type))) + geom_point() + xlab("Log10-transformed Square Footage") + ylab("Log10-transformed Greenhouse Emissions") + ggtitle("Greenhouse Emissions & Building Square Footage") + theme(plot.title = element_text(size=14, face="bold")) + scale_color_manual(values=cbm_palette) + facet_grid(.~year_range)
```

Now, let's look at another metric of energy/resource use in the city. <span style="color:red">*Make a histogram of the water_use variable, applying any transformations you think might be useful* </span>

```{r}
ggplot2::ggplot(energy, aes(x=water_use)) + geom_histogram(bins=100, fill=RColorBrewer::brewer.pal(3, "Set1")[2], color="black") + ggtitle("Histogram of water use") + scale_fill_brewer(palette="Set1")+ theme(plot.title = element_text(size=14, face="bold"))

ggplot2::ggplot(energy, aes(x=log10(water_use))) + geom_histogram(bins=100, fill=RColorBrewer::brewer.pal(3, "Set1")[2], color="black") + ggtitle("Histogram of water use(log10-transformed)") + scale_fill_brewer(palette="Set1")+ theme(plot.title = element_text(size=14, face="bold"))
```

Now, let's compare average water usage by zipcode (postal_code) using side-by-side boxplots. **geom_boxplot()** is the extension of ggplot you'll need for this. Try to color the boxplots by zip code. **Hint**: You may have to convert **postal_code** to a factor from an integer variable. 

```{r}
ggplot2::ggplot(energy, aes(x=as.factor(postal_code), y=log10(water_use), color=as.factor(postal_code))) + geom_boxplot() + ggtitle("Log-transformed Water Use By Zipcode") + scale_fill_brewer(palette="Set1")+ theme(plot.title = element_text(size=14, face="bold"))

```

Finally, let's practice making stacked bar charts. In this step, we'll look at how many observations of each **property_type** we have. To make things a little more interesting, we'll fill the bars for each type by the zipcode the buildings are found in. You can make a bar chart using geom_bar(), and can make this a "stacked" barchart using  **aes(fill= <variable_to_fill_by>)**

```{r}
ggplot2::ggplot(energy, aes(x=property_type, fill=as.factor(postal_code))) + geom_bar() + ggtitle("# of Properties Per Category") + theme(plot.title = element_text(size=14, face="bold")) + xlab("Property Type")  + labs(fill = "Zip Code") + theme(axis.text.x = element_text(angle = 300))

```

```


