---
title: "R Workshop: Session 1"
output: html_document
---

In this section, we will be covering data types, including

* Vectors (numeric logical, and character)
* Lists
* Matrices
* Data Frames
* Factors

Some of these may be familiar from other programming languages, while others are more R-specific


#Vectors

Contain elements of the same type

Building a vector: use the "concatenate" or "combine" function c()

``` {r eval = TRUE}
vector1 <- c(1, 5, 8.4, -6, 2)                  #Numeric
vector2 <- c(TRUE, TRUE, FALSE, NA, FALSE)      #Logical
vector3 <- c("first", "second", "third", "fourth", "fifth")        #Character
```

Vector1 contains integers and floats, but they are all considered of the numeric type by R

* R typically forces ints to floats anyway

Vector3 contains what we would usually consider "strings" in C++ or Java, but R considers them all part of the character class

If you're ever unsure of what type of vector you have, use the class() function

``` {r eval = TRUE}
class(vector1)
```

Accessing vector elements - Indexing

Refer to element(s) of a vector using subscripts enclosed in []

NOTE: R uses 1-indexing

``` {r eval = TRUE}
vector1[3]
vector1[c(2,4)]
```

#Lists

Contain objects that can be of different types, including other lists/vectors; generic vector 

Building a vector: use the list() function

``` {r eval = TRUE}
x <- list(vector1, vector2, vector3, 8)
x
```

Working with Lists - Slicing vs. Referencing 

Slicing, [] - Retrieve a list object or "slice" of the original list

Reference, [[]] - Retrieve a list element, which can them be modified directly

``` {r eval = TRUE}
slice <- x[2]
class(slice)
slice

ref <- x[[2]]
class(ref)
ref
```

#Matrices

2-dimensional (rectangular) collection of data of a single type, with all columns of equal length

Made using the matrix() function

R TIP: type ?[function name] to bring up the manual page for that function

```{r eval = FALSE}
?matrix
```

Based on the manual page, we can make a matrix called 'mymatrix' with the following parameters and then view it:

``` {r eval = TRUE}
data <- c(2, 4, 3, 1, 5, 7)
rnames <- c("row1", "row2")
cnames <- c("col1", "col2", "col3")
mymatrix <- matrix(data = data, nrow = 2, ncol = 3, dimnames = list(rnames, cnames))
mymatrix
```

Now that we have a matrix, we can index by position or name using the format [row,col]

``` {r eval = TRUE}
mymatrix[2,3]
mymatrix["row2", "col3"]
mymatrix[1,]
mymatrix[,2]
```

#Data Frames

2-dimensional collection of data where different columns can contain different data types, generalized matrix

Made using the data.frame() function

``` {r eval = TRUE}
df <- data.frame(vector1, vector2, vector3)
df
```

We can then name the columns

``` {r eval = TRUE}
names(df) <- c("Numbers", "Passed?", "Order")
df
```

#R Quirk: Factors

R explanation : Factors are a data type used to represent categorical data. They look like characters/strings but are actually stored as integers, so they can't always be treated as strings (Ex. "Female" stored as 1, "Male" stored as 2)

Bottom line: if you have a code snippet that you think should be working but isn't, check that the variable classes are what you expect

Let's try to convert number to factors and then back again

``` {r eval = TRUE}
n <- c(3.4, 1.2, 5)   #numeric vector
n
f <- factor(n)        #converting the numbers into factors
f
as.numeric(f)         #converting the factors back to numbers
```

Why did this happen? R stores your data (numeric or not) as integers, regardless of the order you entered them

To see the actual data you saved as factors, use the levels() function

``` {r eval = TRUE}
levels(f)
```

# What are functions?

* A function is essentially a piece of resuable code, that normally has a name.

* We have already used several functions today, for example factor() and levels()



* The basic syntax of a function looks like this:

```{r eval=FALSE}
function_name <- function(arg) {
  #do something
}

```

## Let's write a function!
* We will start off by writing our own "mean()" function


```{r eval=TRUE}
mean2 <- function(a_vector){
  
  lenVec <- length(a_vector)
  sumVec <- sum(a_vector)
  meanVec <- sumVec / lenVec
  
}

```

* Time to test our mean function on the vector c(2,5,6,7):

```{r eval=TRUE}
vecMean <- mean2(c(2,5,6,7))
print(vecMean)

 mean(c(2,5,6,7))
```

* We can take a look at any functions we have loaded into our workspace by typing "function_name".

* Let's take a look at one we are already familiar with:


```{r eval=TRUE}
mean2

```

## How do I save a function?
* One of the useful things about functions is that they can be saved and used later.
* By saving the function in a "my_function.R" file you can run these functions just like you would run any built in function

#### WARNING NEED TO KEEP AN EYE ON WHAT YOUR WORKING DIRECTORY IS!


* If you find that you are using the same arguments frequently then you can give your arguments default parameters by using:

```{r eval=FALSE}
my_function <- function(arg=default){
  
  #do something
  
}

```


* Let's write a function that uses a default argument! 
* Write a function called addSumNumbers, that adds two numbers, but by default adds 1 to a number that has no default
* Then run addSumNumbers using 5 and the default.

```{r eval=TRUE}
addSumNumbers <- function(numOne, numTwo=1){
  
 numOne + numTwo
  
}

print(addSumNumbers(5))
```

* You can switch the order of the arguments if you name things. Let's try this using a new function, divideSumNumbers

```{r eval=TRUE}
divideSumNumbers <- function(numOne, numTwo=1){
  
 numOne / numTwo
  
}

print(divideSumNumbers(5))
print(divideSumNumbers(numTwo=1,numOne=5))
```

* R can also autocomplete your argument names.
* Try "print(divideSumNumbers(numT=1,numO=5))"

```{r eval=TRUE}
print(divideSumNumbers(numT=1,numO=5))
```

* I recommend being careful with this functionality, because it can make your code a bit tricky to read.

## Useful Function Tricks
* You can have a function be an argumenent in a second function
* lets give this a shot by writing a function called "evaluate" that has the arguments "a function you want to run"(like mean() or sum()) and a vector of numbers
```{r eval=TRUE}
evaluate <- function(func, dat){
  
  func(dat)
  
}
```

* Now test evalute using by taking the sum of the vector c(2,4,6)

```{r eval=TRUE}

print(evaluate(mean,c(2,4,6)))

```

## Anonymous Functions

* A function does not need to have a name, if you just want to declare a function and not name it, it is called a "anonymous fucntion"

* Here is an example
```{r eval=FALSE}
function(x){x+1}
```

* These can be particularly handy in a loop (Which we will talk about shortly), but for now let's try a few things using our evaluate function.

* use evaluate and an anomyous function to add 1 to the number 5
```{r eval=TRUE}
print(evaluate(function(x){x[1]}, 5))
```

* use evaluate and an anomyous function to return the last item in this vector c(2,4,6)
```{r eval=TRUE}
evaluate(function(x){x[-1]}, c(8, 4, 0))
```

# LOOPS!
* Loops are handy when you want to iterate over the individual units in something.
* For example: entries in a list, columns in a dataframe, or numbers in a vector

#### Disclaimer: Should vectorize when possible, but that is a topic for another day.

# BACK TO LOOPS!

*You saw before that to find the data type for each element in a data vector we had to write out the same command for each element. To perform the same command over each element we can use loops. 

*Basic for loop:

```{r eval=TRUE}
lst = list('text', 1, 'chr', 3.0)

for (x in lst){
  
  print(typeof(x))
}
```


* For Loops are useful, but R has an even more compact way of looping over elements that also have handy features when you are coding. These are the *apply family of functions, apply, sapply, lapply, vapply, and tapply (there are others but we'll focus on these here)

## Apply function
* Apply is great for iterating a function over an object, such as a dataframe or a matrix

* The basic anatomy of an apply command is:
```{r eval=FALSE}
apply(object, direction, function)

```
* Where the direction is either 1 (for iterate over the rows) or 2 for iterate over columns

Let's try this out using the Framingham data set. First, Let's find out what class each column of the dataframe is.

```{r eval=TRUE}
#read in the data
fhs = read.csv("Framingham.dat")

colClass <- apply(fhs, 2, class)

colClass
```

Now, let's try using apply on a single column of our data. Let's use apply to calculate each individual's distance from the mean for their SBP
```{r eval=TRUE}

meanSBP <- mean(fhs$SBP)
SBPdiffMean <- apply(fhs, 1, function(x){as.numeric(x["SBP"]) - meanSBP})

print(meanSBP)
head(SBPdiffMean)
tail(SBPdiffMean)
head(fhs)
tail(fhs)
```

##lappy

* There are variations on apply that are nice for knowing exactly what kind of object will come out the other end a good example is lappy. lappy works a lot like apply, but it returns the output in the form of a list.

* Try using lappy() to look at the classes of the different columns in the Framingham data.

```{r eval=TRUE}

colClassLst <- lapply(fhs, class)

colClassLst
```

## sapply

* sapply is similar to lapply, where the "s" is for "simple". It will attempt to simplify the output of your loop as far as it can, so for example instead of returning a list (lappy) it will return a vector.

*Lets do the same command we did with lappy using sapply

```{r eval=TRUE}

colClassVec <- sapply(fhs, class)
colClassVec

```

* Sapply can be handy for looking at basic stats for multiple columns in a dataframe. As example use sapply to get the range of the last 4 columns of the Framingham data.
* What kind of object do you think it will output?

```{r eval=TRUE}

sapply(fhs[,4:8], range)

```

## vapply

* vapply is again similar to apply, but it requires you to designate what you expect the output object to be, and if the output is not what you expect it will throw an error. This is particularly handy if you are writing a script.

* Let's try to use vapply to get a character vector of the column classes
```{r eval=TRUE}

colClassVec <- vapply(fhs, class, character(1))
colClassVec

```

## tapply

* tapply works similar to table, but allows you to apply a function of each "factor" in the column you are "tabling"

* To try out tapply find the range, median, and mean of the male and female SBPs.
```{r eval=TRUE}

tapply(fhs$SBP, fhs$SEX, summary )

```
