---
title: "R_workshop_preamble"
author: "Binglan Li"
date: "8/13/2019"
output: html_document
---

# Preamble

## 1. Installing R Packages

Run the following chunk of codes to install necessary packages for this workshop. These packages can be useful for statistical analysis or data visualization for your future class work or research.

```{r install packages, eval=FALSE, include=FALSE}
# the command to install a R package
install.packages("ggplot2", repos = "https://cran.rstudio.com")
install.packages("dplyr", repos = "https://cran.rstudio.com")
install.packages("RColorBrewer", repos = "https://cran.rstudio.com")

# ggplot2 and dplyr are parts of a larger R package called 'tidyverse'
# you can use the following code just install all packages in 'tidyverse'
#install.packages("tidyverse")

# if you cannot install certain packages using the above codes, try the following
#devtools::install_github("tidyverse/ggplot2")
#devtools::install_github("tidyverse/dplyr")
```

## 2. Learn to Load R Packages

Remember that successfully installing a R package does not mean that it will be automatically loaded to your R environment afterwards. Everytime you newly install a R package or open Rstudio, you need to load the R packages by yourself using **libarary()**. Run the following chunk of codes to load R packages.

```{r load packages, eval=FALSE, include=FALSE}
# now, actual commands to load library
library(ggplot2)
library(dplyr)
library(RColorBrewer)

# now use the command to see if you have successfully load ggplot2 in your environment
search() 
```

## 3. Helpful functions or features in Rstudio

  1. Automatic filling using the 'Tab' key. 

```{r helpful function, eval=FALSE, include=FALSE}
# automatic filling
# try to find the mean() function, which calculates the mean of input vectors, using the 'tab' key immediate folloing 'me'
me
```

  2. **?function_name** to look up description and manual for a R function

```{r helpful function, echo=TRUE}
# manual lookup
?mean
```

  3. **Command + Enter** in Mac and **Control + Enter** in Windows to execute the line of source code where the cursor currently resides

```{r helpful function, echo=TRUE}
# try to execute the following code with the shortcut
mean(c(1,2,3))
```

  4. **<-** and **=** are used interchangeably in R to assign values to a variable

```{r helpful function, echo=TRUE}
# assigning values using arror head
mean1 <- mean(c(1,2,3))
mean1
# assigning value using the classic equal sign
mean2 = mean(c(1,2,3))
mean2
```


## Reference and Helpful Materials

#### Online R Textbook
* R for Data Science (https://r4ds.had.co.nz/)
* Advanced R (https://adv-r.hadley.nz/)

#### Statistical Analysis in R
* An Introduction to Statistical Learning: With Applications in R
* Applied linear statistical models

#### Visualization in R
* R cheat sheet (https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)
* ggplot2 cheat sheet (https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
* Google ggplot2 reference to to go the following website, https://ggplot2.tidyverse.org/reference/index.html
* Stack Overflow with keywords, “R” or “ggplot2”
* R for Data Science available at https://r4ds.had.co.nz/
* The R Graphics Cookbook by Winston Chang

